### 2021-08-04 paper/localpsf.tex
notation in appendix fixes

Since we use finite elements to discretize the problem, the $i^\text{th}$ column of $\widetilde{\mathbf{A}}$ corresponds to the nodal point in $\mathbb{R}^d$ associated with the $i^\text{th}$ finite element basis function. The columns of the matrix thus correspond to a point cloud in $\mathbb{R}^d$. We split this point cloud into two equally sized \emph{child} point clouds, using a hyperplane which is perpendicular to the coordinate axis direction in which the point cloud is widest (e.g., either the $x$, $y$, or $z$ axis in $3$D). The two child point clouds are split in the same way. This splitting process repeats until the point clouds have less than a preset number of points (we use $32$ points). This hierarchical partitioning of the point cloud into smaller and smaller point clouds corresponds to a hierarchical partitioning of the columns of the matrix into smaller and smaller \emph{clusters} of columns. This hierarchical partitioning of the columns forms a binary tree, which is called the column cluster tree. The root of the tree is the set of all columns, and the leaves of the tree are clusters of columns that are not subdivided any further.

@@

After reordering the rows and columns of $\widetilde{\mathbf{A}}$ via the depth-first ordering described above, we partition the reordered version of $\widetilde{\mathbf{A}}$ into a tree of $2 \times 2$ block matrices recursively. We use a geometric admissibility condition (discussed below) to decide which blocks to subdivide, and use the cluster trees for the rows and columns to determine how to subdivide those blocks. For the first stage of subdivision, let $r_1$ and $r_2$ be the children row clusters for the root of the row cluster tree, let $c_1$ and $c_2$ be the children column clusters for the root of the column cluster tree. The matrix $\widetilde{\mathbf{A}}$ is partitioned into blocks as follows:
\begin{equation*}
\begin{bmatrix}
\widetilde{\mathbf{A}}_{11} & \widetilde{\mathbf{A}}_{12} \\
\widetilde{\mathbf{A}}_{21} & \widetilde{\mathbf{A}}_{22},
\end{bmatrix}
\end{equation*}
where $\widetilde{\mathbf{A}}_{11}$ denotes the block of $\widetilde{\mathbf{A}}$ with rows $r_1$ and columns $c_1$, $\widetilde{\mathbf{A}}_{12}$ denotes the block of $\widetilde{\mathbf{A}}$ with rows $r_1$ and columns $c_2$, and so on for $\widetilde{\mathbf{A}}_{21}$ and $\widetilde{\mathbf{A}}_{22}$.

We now loop through the four blocks, $\widetilde{\mathbf{A}}_{11}$, $\widetilde{\mathbf{A}}_{12}$, $\widetilde{\mathbf{A}}_{21}$, and $\widetilde{\mathbf{A}}_{22}$, and decide which blocks should be subdivided further. For the purpose of explanation, consider $\widetilde{\mathbf{A}}_{12}$. If
\begin{equation}
\label{eq:weak_admissibility_cond}
\dist\left(r_1, c_2\right) \ge \eta \min\left(\diam\left(r_1\right), \diam\left(c_2\right)\right),
\end{equation}
then we mark $\widetilde{\mathbf{A}}_{12}$ as \emph{admissible} (expected to be low rank) and leave it alone. Here $\dist\left(r_1, c_2\right)$ is the Euclidean distance between the axis-aligned bounding box for the point cloud associated with the row cluster $r_1$, and the axis aligned bounding box for the point cloud associated with the column cluster $c_2$. The quantity $\diam\left(r_1\right)$ is the diameter of the axis aligned bounding box for the point cloud associated with the row cluster $r_1$, and $\diam\left(c_2\right)$ is the analogous diameter associated with the column cluster $c_2$. The quantity $\eta$ is a scalar constant; we use $\eta=2.0$. Basically, if the point clouds associated with $r_1$ and $c_2$ are far away from each other relative to their diameters, then we expect that the corresponding block of the matrix will be low rank. This process is repeated for the other blocks to determine which blocks are admissible and which are not. For us, the diagonal blocks $\widetilde{\mathbf{A}}_{11}$ and $\widetilde{\mathbf{A}}_{22}$ are not admissible because the distance between a point cloud and itself is zero.

Next, we subdivide all blocks that are not admissible and are larger than a predetermined size (we use size $32 \times 32$), using the same process that was used to subdivide $\mathbf{\widetilde{A}}$. But now we subdivide a block based on the two child row clusters and two child column clusters for the rows and columns of that block. This subdivision process continues recursively until all blocks are either admissible, or smaller than the predetermined size mentioned above. The resulting hierarchical partitioning of matrix blocks forms a tree, which is called the \emph{block cluster tree}. The root of the tree is the whole matrix, internal nodes in the tree are blocks that are subdivided, and the leaves of the tree are blocks that are either expected to be low rank, or are small.

@@

Once the block cluster tree has been constructed, low rank approximations of the admissible (low rank) blocks are formed using the adaptive cross method \cite{ACA}. Let $X \in \mathbb{R}^{m \times m}$ be an admissible block of $\widetilde{\mathbf{A}}$. The idea of the adaptive cross method is to form a low rank approximation of $X$ by sampling a small number of rows and columns of $X$. 
\begin{itemize}
	\item Let $C \in \mathbb{R}^{m \times r}$ be a matrix consisting of a subset of $r$ columns of $X$, such that the span of the columns in $C$ approximates the column space of $X$. 
	\item Let $R\in\mathbb{R}^{r \times m}$ be a subset of the rows of $X$, such that the span of the rows in $R$ approximates the row space of $X$.
	\item Let $U \in \mathbb{R}^{r \times r}$ be the block of $X$ consisting of the intersection of the rows from $R$ with the columns from $C$.
\end{itemize}
Then it is well-established that
\begin{equation}
\label{eq:CUR}
X \approx C U^+ R,
\end{equation}
where $U^+$ is the pseudoinverse of $U$ \cite{CUR}. The quality of approximation \eqref{eq:CUR} depends on how well the columns of $C$ approximate the column space of $X$, and how well the rows of $R$ approximate the row space of $X$.

@@

We use the adaptive cross method to form low rank approximations for each admissible block of $\widetilde{\mathbf{A}}$. We directly compute all entries of the small dense blocks of $\widetilde{\mathbf{A}}$ that are not admissible. This process requires us to compute $O(r^2 N \log N)$ entries of $\widetilde{\mathbf{A}}$, which is relatively cheap compared to the matrix-vector products with $\mathbf{A}$ that are required to form the product-convolution approximation $\widetilde{\mathbf{A}}$.

>>>>

Since we use finite elements to discretize the problem, the $i^\text{th}$ column of $\AkerPcMat$ corresponds to the nodal point in $\mathbb{R}^d$ associated with the $i^\text{th}$ finite element basis function. The columns of the matrix thus correspond to a point cloud in $\mathbb{R}^d$. We split this point cloud into two equally sized \emph{child} point clouds, using a hyperplane which is perpendicular to the coordinate axis direction in which the point cloud is widest (e.g., either the $x$, $y$, or $z$ axis in $3$D). The two child point clouds are split in the same way. This splitting process repeats until the point clouds have less than a preset number of points (we use $32$ points). This hierarchical partitioning of the point cloud into smaller and smaller point clouds corresponds to a hierarchical partitioning of the columns of the matrix into smaller and smaller \emph{clusters} of columns. This hierarchical partitioning of the columns forms a binary tree, which is called the column cluster tree. The root of the tree is the set of all columns, and the leaves of the tree are clusters of columns that are not subdivided any further.

@@

After reordering the rows and columns of $\AkerPcMat$ via the depth-first ordering described above, we partition the reordered version of $\AkerPcMat$ into a tree of $2 \times 2$ block matrices recursively. We use a geometric admissibility condition (discussed below) to decide which blocks to subdivide, and use the cluster trees for the rows and columns to determine how to subdivide those blocks. For the first stage of subdivision, let $r_1$ and $r_2$ be the children row clusters for the root of the row cluster tree, let $c_1$ and $c_2$ be the children column clusters for the root of the column cluster tree. The matrix $\AkerPcMat$ is partitioned into blocks as follows:
\begin{equation*}
\begin{bmatrix}
\left(\AkerPcMat\right)_{11} & \left(\AkerPcMat\right)_{12} \\
\left(\AkerPcMat\right)_{21} & \left(\AkerPcMat\right)_{22},
\end{bmatrix}
\end{equation*}
where $\left(\AkerPcMat\right)_{11}$ denotes the block of $\AkerPcMat$ with rows $r_1$ and columns $c_1$, $\left(\AkerPcMat\right)_{12}$ denotes the block of $\AkerPcMat$ with rows $r_1$ and columns $c_2$, and so on for $\left(\AkerPcMat\right)_{21}$ and $\left(\AkerPcMat\right)_{22}$.

We now loop through the four blocks, $\left(\AkerPcMat\right)_{11}$, $\left(\AkerPcMat\right)_{12}$, $\left(\AkerPcMat\right)_{21}$, and $\left(\AkerPcMat\right)_{22}$, and decide which blocks should be subdivided further. For the purpose of explanation, consider $\left(\AkerPcMat\right)_{12}$. If
\begin{equation}
\label{eq:weak_admissibility_cond}
\dist\left(r_1, c_2\right) \ge \eta \min\left(\diam\left(r_1\right), \diam\left(c_2\right)\right),
\end{equation}
then we mark $\left(\AkerPcMat\right)_{12}$ as \emph{admissible} (expected to be low rank) and leave it alone. Here $\dist\left(r_1, c_2\right)$ is the Euclidean distance between the axis-aligned bounding box for the point cloud associated with the row cluster $r_1$, and the axis aligned bounding box for the point cloud associated with the column cluster $c_2$. The quantity $\diam\left(r_1\right)$ is the diameter of the axis aligned bounding box for the point cloud associated with the row cluster $r_1$, and $\diam\left(c_2\right)$ is the analogous diameter associated with the column cluster $c_2$. The quantity $\eta$ is a scalar constant; we use $\eta=2.0$. Basically, if the point clouds associated with $r_1$ and $c_2$ are far away from each other relative to their diameters, then we expect that the corresponding block of the matrix will be low rank. This process is repeated for the other blocks to determine which blocks are admissible and which are not. For us, the diagonal blocks $\left(\AkerPcMat\right)_{11}$ and $\left(\AkerPcMat\right)_{22}$ are not admissible because the distance between a point cloud and itself is zero.

Next, we subdivide all blocks that are not admissible and are larger than a predetermined size (we use size $32 \times 32$), using the same process that was used to subdivide $\AkerPcMat$. But now we subdivide a block based on the two child row clusters and two child column clusters for the rows and columns of that block. This subdivision process continues recursively until all blocks are either admissible, or smaller than the predetermined size mentioned above. The resulting hierarchical partitioning of matrix blocks forms a tree, which is called the \emph{block cluster tree}. The root of the tree is the whole matrix, internal nodes in the tree are blocks that are subdivided, and the leaves of the tree are blocks that are either expected to be low rank, or are small.

@@

Once the block cluster tree has been constructed, low rank approximations of the admissible (low rank) blocks are formed using the adaptive cross method \cite{ACA}. Let $X \in \mathbb{R}^{m \times m}$ be an admissible block of $\AkerPcMat$. The idea of the adaptive cross method is to form a low rank approximation of $X$ by sampling a small number of rows and columns of $X$. 
\begin{itemize}
	\item Let $C \in \mathbb{R}^{m \times r}$ be a matrix consisting of a subset of $r$ columns of $X$, such that the span of the columns in $C$ approximates the column space of $X$. 
	\item Let $R\in\mathbb{R}^{r \times m}$ be a subset of the rows of $X$, such that the span of the rows in $R$ approximates the row space of $X$.
	\item Let $U \in \mathbb{R}^{r \times r}$ be the block of $X$ consisting of the intersection of the rows from $R$ with the columns from $C$.
\end{itemize}
Then it is well-established that
\begin{equation}
\label{eq:CUR}
X \approx C U^+ R,
\end{equation}
where $U^+$ is the pseudoinverse of $U$ \cite{CUR}. The quality of approximation \eqref{eq:CUR} depends on how well the columns of $C$ approximate the column space of $X$, and how well the rows of $R$ approximate the row space of $X$.

@@

We use the adaptive cross method to form low rank approximations for each admissible block of $\AkerPcMat$. We directly compute all entries of the small dense blocks of $\AkerPcMat$ that are not admissible. This process requires us to compute $O(r^2 N \log N)$ entries of $\AkerPcMat$, which is relatively cheap compared to the operator actions of $\Aop$ that are required to form the product-convolution approximation.
